---
title: "SI compartmental model with imperfect vaccination"
author: "hermina ghenu"
date: "2025-03-26"
output: html_document
runtime: shiny
---

# Goal

I am building the model presented in [Kribs-Zaleta & Velasco-Hernández 2000](https://www.sciencedirect.com/science/article/pii/S0025556400000031). I used the infectious_diseases from [the EcoEvoApps project](https://gitlab.com/ecoevoapps/ecoevoapps) as my starting point.

My goal is to start playing around with an R Shiny app of a compartmental model in order to design a coherent tutorial that can help students learn. In particular, I want to see if it's possible to teach students how to make their own extensions to an existing model.

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
library(deSolve)
library(patchwork)
library(kableExtra)
library(shiny)
library(ecoevoapps) # now only used for the theme
theme_set(ecoevoapps::theme_apps())
```


# Description of the Biological Problem / Question

# Explanation of the Model

## Model Assumptions

[Insert text here]

## Define variables & parameters

Variables:

```{r vars SVI, echo=F}
# different columns of the table
vars <- c("$S$",
          "$I$",
          "$V$",
          "$T$")
vars_desc <- c("Population fraction of susceptible individuals",
               "Population fraction of infectious individuals",
               "Population fraction of vaccinated individuals",
               "Time"
                 )
vars_units <- c("none (susceptible individuals / total individuals)",
                "none (infected individuals / total individuals)",
                "none (vaccinated individuals / total individuals)",
                "days, weeks, years, etc.")

# combine columns and render the table nicely
var_df <- data.frame(vars, vars_desc, vars_units)
kable(x = var_df, format = "html", 
      col.names = c("Variable", "Description", "Example unit")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
# clean up
rm(vars, vars_desc, vars_units, var_df)
```  

Parameters:

```{r pars SVI, echo=F}
# different columns of the table
pars      <- c("$\\beta$", 
               "$\\gamma$ (c in 2000 paper)",
               "$\\theta$",
               "$\\sigma$",
               "$\\phi$")
pars_desc <- c("Transmission rate",
               "Recovery rate",
               "Vaccine waning rate",
               "Modifier of trasmission rate thanks to vaccination",
               "Vaccination rate"
                 )
min_max <- c("[0, +Inf)",
             "[0, +Inf)",
             "[0, +Inf)",
             "[0, 1]",
             "[0, +Inf)")

# combine columns and render the table nicely
param_df <- data.frame(pars, pars_desc, min_max)
kable(x = param_df, format = "html", 
      col.names = c("Parameter", "Description", "Range")) %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed"),
                position = "center")
# clean up
rm(pars, pars_desc, min_max, param_df)
```  


## Qualitative description

Let's create another version of this plot that is native to ggplot. This way it can later become an interactive part of the shiny app:

```{r schematic_diag, echo=FALSE}
# a function to define the corners of the 3 rectangles
make_rectangles <- function(S_scale=1, # looks good: min 0.1 and max 3
                            I_scale=1, # looks good: min 0.1 and max 3
                            V_scale=1, # looks good: min 0.1 and max 3
                            beta_scale=0.5, # transmission (S -> I): min 0.1, max 2.2
                            gamma_scale=0.5, # recovery (I -> S): min 0.1, max 2.2
                            phi_scale=0.5, # vaccination policy (S -> V): min 0.1, max 2.2
                            theta_scale=0.5, # vaccination waning (V -> S): min 0.1, max 2.2
                            sigma_scale=0.5 # modifier of transmission rate (V -> I): min 0.1, max 2.2
                            ){
  # check if vaccination is enabled
  vaccine <- (V_scale | phi_scale) # at least one must be non-zero
  
  # define the standard ratio of the basic rectangle
  rect_HALFlength <- 1.7/2
  rect_HALFheight <- 0.4/2
  # define the fixed center point of each rectangle
  S_center <- c(x=0.85, y=2.8)
  I_center <- c(x=4.15, y=2.8)
  V_center <- c(x=2.15, y=1.7)
  
  # update the size of each rectangle based on its scale
  S_len <- S_scale * rect_HALFlength
  S_hei <- S_scale * rect_HALFheight
  I_len <- I_scale * rect_HALFlength
  I_hei <- I_scale * rect_HALFheight
  V_len <- V_scale * rect_HALFlength
  V_hei <- V_scale * rect_HALFheight
  
  # Create a data.frame to define the rectangles and text
  rect_coords <- data.frame(
    x_center = c(S_center["x"], I_center["x"]),
    y_center = c(S_center["y"], I_center["y"]),
    left = c(S_center["x"]-S_len, I_center["x"]-I_len),
    right = c(S_center["x"]+S_len, I_center["x"]+I_len),
    bottom = c(S_center["y"]-S_hei, I_center["y"]-I_hei),
    top = c(S_center["y"]+S_hei, I_center["y"]+I_hei)
  )
  # name the rows by the compartment
  row.names(rect_coords) = c("Susceptible", "Infected")
  # a named vector for the colours of the compartment rectangles
  rect_colours <- c(Susceptible = "#440154FF", Infected = "#FDE725FF")
  
  # if vaccination is enabled, add the third "vaccinated" rectangle
  if (vaccine) {
    rect_coords <- rbind(rect_coords,
                         c(V_center["x"], V_center["y"],
                           V_center["x"]-V_len, V_center["x"]+V_len,
                           V_center["y"]-V_hei, V_center["y"]+V_hei))
    row.names(rect_coords)[3] <- "Vaccinated"
    rect_colours <- c(rect_colours, Vaccinated =  "#21908CFF")
  }
  
  # begin making the plot. Put the rectangles on the lowest layer
  output <- ggplot() +
            # Add rectangles
            geom_rect(data = rect_coords,
                      aes(xmin = left, xmax = right, ymin = bottom, ymax = top), 
                      fill = rect_colours, color = "black", alpha=0.6) +
            # Add text labels
            geom_text(data = rect_coords,
                      aes(x = x_center, y = y_center, label = row.names(rect_coords)), 
                      size = 5)
  
  # if transmission > 0, add the transmission rate arrow
  if (beta_scale > 0){
    output <- output +
            # plot the transmission arrow
              geom_curve(aes(x = rect_coords["Susceptible", "x_center"]+0.8,
                             xend = rect_coords["Infected", "x_center"]-0.5,
                             y = rect_coords["Susceptible", "y_center"]+0.05,
                             yend = rect_coords["Infected", "y_center"]+0.05),
                         curvature = -0.5, lineend = "butt",
                         arrow = arrow(length = unit(0.04, "npc")),
              # weight of arrow depends on beta parameter
                         linewidth=beta_scale) +
            # add transmission label
              geom_text(aes(x = 0.05 + rect_coords["Susceptible", "x_center"] + (rect_coords["Infected", "x_center"] - rect_coords["Susceptible", "x_center"])/2,
                            y = rect_coords["Susceptible", "y_center"]+0.4,
                        label = "beta"),
                        size = 5, parse=TRUE)
  }
  
  # if recovery > 0, add the recovery rate arrow
  if (gamma_scale > 0) {
    output <- output +
            # the recovery arrow
              geom_curve(aes(xend = rect_coords["Susceptible", "x_center"]+0.75,
                             x = rect_coords["Infected", "x_center"]-0.55,
                             yend = rect_coords["Susceptible", "y_center"]-0.05,
                             y = rect_coords["Infected", "y_center"]-0.05),
                         curvature = -0.5, lineend = "butt",
                         arrow = arrow(length = unit(0.04, "npc")),
              # weight of arrow depends on gamma parameter
                         linewidth=gamma_scale) +
            # add recovery label
              geom_text(aes(x = 0.05 + rect_coords["Susceptible", "x_center"] + (rect_coords["Infected", "x_center"] - rect_coords["Susceptible", "x_center"])/2,
                            y = rect_coords["Susceptible", "y_center"]-0.38,
                        label = "gamma"),
                        size = 5, parse=TRUE)
  }
  
  # if vaccination policy > 0, add the vaccination rate arrow
  if (phi_scale > 0){
    output <- output +
            # define the vaccination rate arrow
              geom_curve(aes(x = rect_coords["Susceptible", "x_center"]+0.15,
                             xend = rect_coords["Vaccinated", "x_center"]-0.5,
                             y = rect_coords["Susceptible", "y_center"]-0.11,
                             yend = rect_coords["Vaccinated", "y_center"]+0.07),
                         curvature = -0.25, lineend = "butt",
                         arrow = arrow(length = unit(0.04, "npc")),
            # weight of arrow depends on phi parameter
                         linewidth=phi_scale) +
            # add vaccination rate label
              geom_text(aes(x=1.75, y=2.3, label = "phi"),
                        size = 5, parse=TRUE)
  }
  
  # if vaccine can wane, add the waning rate arrow
  if (vaccine & (theta_scale > 0)) { # to prevent extraneous arrows or labels, need to check that vaccine is also TRUE
    output <- output +
            # define the waning rate arrow
              geom_curve(aes(xend = rect_coords["Susceptible", "x_center"]-0.05,
                             x = rect_coords["Vaccinated", "x_center"]-0.7,
                             yend = rect_coords["Susceptible", "y_center"]-0.11,
                             y = rect_coords["Vaccinated", "y_center"]+0.07),
                         curvature = -0.25, lineend = "butt",
                         arrow = arrow(length = unit(0.04, "npc")),
            # weight of arrow depends on theta parameter
                         linewidth=theta_scale) +
            # add waning rate label
              geom_text(aes(x=0.6, y=2.3, label = "theta"),
                        size = 5, parse=TRUE)
  }
  
  #if vaccine is not perfectly effective, add modified transmission arrows
  if (vaccine & (sigma_scale > 0)){ # to prevent extraneous arrows or labels, need to check that vaccine is also TRUE
    output <- output +
          # finally define the vaccine-modified transmission rate arrow
              geom_curve(aes(x = rect_coords["Vaccinated", "x_center"]+0.7,
                             xend = rect_coords["Infected", "x_center"]-0.05,
                             y = rect_coords["Vaccinated", "y_center"],
                             yend = rect_coords["Susceptible", "y_center"]-0.11),
                         curvature = 0.4, lineend = "butt",
                         arrow = arrow(length = unit(0.04, "npc")),
              # weight of arrow depends on sigma*beta parameters
                         linewidth=sigma_scale) +
              # add vaccination rate label
              geom_text(aes(x=4.27, y=2.05, label = "sigma * beta"),
                        size = 5, parse=TRUE)
  }
  
  # finally, set limits and theme
  if (vaccine) {
    output <- output +
              xlim(min(rect_coords$left), max(rect_coords$right)) +
              ylim(min(rect_coords$bottom), max(c(rect_coords$top, 3.2))) +
              theme_void()
  } else {
    output <- output +
              xlim(min(rect_coords$left), max(rect_coords$right)) +
              ylim(min(rect_coords$bottom, 2), max(c(rect_coords$top, 3.5))) + # y axis has to be artificially forced to be bigger than min/max values otherwise the boxes and arrows can look weird
              theme_void()
  }
  
  return(output)
}

# plot the schematic
print(make_rectangles())

# plot again without vaccinated
#print(make_rectangles(S_scale=1, I_scale=1, V_scale=0, phi_scale=0))
```

## Quantitative description

Description of the model modified from pg. 212-3 [Blackwood & Childs 2018](https://lettersinbiomath.org/manuscript/index.php/lib/article/view/77):

Assuming that the fraction of individuals in the susceptible class is given by S, vaccinated class is given by V, and infectious class is given by I

\[
\begin{align}
\frac{dS}{dt} &= \theta V + \gamma I - \beta I S - \phi S,\\
\frac{dI}{dt} &= \beta I S + \sigma\beta I V - \gamma I,
\frac{dV}{dt} &= \phi S - \sigma\beta I V - \theta V,\\
\end{align}
\]

where the total population size $N=S+V+I=1$, $\lambda(I)=\beta(I/N)=\beta I$ is the force of infection to the susceptible class, $\lambda_V (I)=\sigma\beta(I/N)=\sigma\beta I$ is the force of infection to the vaccinated class, $\beta$ is the transmission rate, $\theta$ is the vaccine waning rate, $\sigma$ is a modifier of the transmission rate such that vaccinated individuals produce fewer infections ($\sigma<1$) than susceptible individuals, $\gamma$ is the recovery rate, and $\phi$ is the vaccination rate. There are no births or deaths in this population: this is sometimes a reasonable assumption. When the rate of waning of immunity from the vaccine exceeds that of the recovery rate, it is possible to have a stable endemic equilibrium while $R_0 < 1$.

Show a bit of the code so that students can get an idea of what it looks like:

```{r SVI_model_show}
# define the SVI model
#' SVI model with vital rates
#' @param time vector of time units over which to run model
#' @param init initial population size of population
#' @param params beta (transmission rate), gamma (recovery rate),
#' theta (vaccine waning rate),
#' sigma (ability of vaccinated to produce new infections),
#' and phi (vaccination rate).
#' @keywords internal
SVI <- function(time,init,params) {
  with (as.list(c(time,init,params)), {
    # description of parameters:
    # beta (transmission rate)
    # gamma (recovery rate)
    # theta (vaccine waning rate)
    # sigma (ability of vaccinated to produce new infections)
    # phi (vaccination rate)
    # S + V + I = 1
    dS_dt = theta*V + gamma*I - beta*I*S - phi*S
    dV_dt = phi*S - (sigma*beta)*I*V - theta*V
    dI_dt = beta*S*I + (sigma*beta)*V*I - gamma*I
    return(list(c(dS = dS_dt, dV = dV_dt, dI = dI_dt)))
  })
}
```

```{r SVI_model_hide, echo=FALSE}
# a function to run forward time simulations of the SVI model
run_SVI_model <- function(time, init, params) {
  # first check the analytical results from the paper
  soln <- get_bifurcation(init = init, params = params)
  # check for fully neutral system
  if (soln$behav == "neutral equilibrium: system stays the same as initial.") {
    # it's trivial to simulate the neutral dynamics:
    data.frame(time = time, S = unname(init["S"]), V = unname(init["V"]), I = unname(init["I"]))
  } else {
    # solve the system of ODE's for the requested time points:
    data.frame(ode(func = SVI, y = init, parms = params, times = time, method="ode23"))
  }
  # ode23 is better than others (adams is the next best but definitely worse)
  # these both still have some issues at params = c(b=3.82, c=1, w=0, m=0.04, p=0.196) and init = c(S=0.9, V=0, I=0.1). This problem is specifically when w=0 then the I category disappears after ~ 50 time units (but it should be a flat line at 0)
  # another issue is when there's no movement into the vaccinated compartment but there is movement out of it, especially once the movement out of that compartment gets fairly quick. For example: params = c(b=2.27, c=0.79, w=1.46, m=0.48, p=0) and init = c(S=0.8, V=0.1, I=0.1)
  # (ode23 could be faster ... but it's not as bad as ode45, which times out for large b)
  # euler and rk4 are just awful
}

#' Plot solutions forward in time for SVI model
#' @param sim_df simulated data frame generated from run_SVI_model()
#' @import ggplot2
#' @seealso [plot_infectiousdisease_time()] this code is based off of that
#' @examples
#' # Run SVI model
#' params_vec <- c(mu = 0.01, beta = 3.27, gamma = 0.5, theta = 0.01,
#'                 sigma = 0.02, phi = 100)
#' params_vec2 <- c(mu = 0.01, beta = 4, gamma = 0.1, theta = 0.01,
#'                 sigma = 0.2, phi = 100)
#' init_vec <- c(S = 1 - 10^-4, V = 0, I = 10^-4)
#' time_vec <- seq(0, 100, 0.1)
#' svi_out <- run_SVI_model(time = time_vec, init = init_vec, params = params_vec)
#' plot_SVI_time(svi_out, params = params_vec)
#' @export
plot_SVI_time <- function(sim_df, params, title_text=NA) {

  # To suppress CMD Check
  R <- S <- E <- D <- time <- Group <- value <- NULL
  
  # check if vaccination is enabled
  vaccine <- (sim_df$V[1] | params["phi"]) # at least one must be non-zero
  
  # change the data.frame from wide format to long format for plotting
  sim_df_long <- sim_df %>% rename(Susceptible = S, Vaccinated = V, Infected = I) %>%
                  pivot_longer(cols = !time, names_to = "Group") %>%
                    mutate(Group = factor(Group, levels = c("Susceptible", "Vaccinated", "Infected")))
  
  # if vaccinated compartment is fixed at zero for all 
  if (!vaccine) {
    # remove the redundant 0's
    sim_df_long <- sim_df_long %>% filter(Group != "Vaccinated")
    sim_df_long$Group <- droplevels(sim_df_long$Group)
  }
  
  #plot for finite time
  finite_plot <- ggplot(sim_df_long) +
                  # this could be a crappy way to deal with numerical problems...
                  #geom_smooth(aes(x = time, y = value, colour = Group),
                   #           method = "loess", se=FALSE, linewidth = 2, alpha = 0.8) +              
                  geom_line(aes(x = time, y = value, color = Group), linewidth = 2.5, alpha=0.8) +
                  scale_colour_viridis_d() +
                  ylab("Population fraction") +
                  scale_y_continuous(limits = c(-0.01, 1.01), expand = c(0, 0)) + # important to set expand to 0,0 otherwise there's extra padding that makes it look like things aren't going to 0 or 1
    # the problem is that sometimes the lines disappear bc they are beyond the grid range
                  theme_apps() + theme(text = element_text(size=18), legend.position="none")
  
  # get the state of the system as time -> +Inf by using the analytic results from the paper
  soln <- get_bifurcation(init = unlist(sim_df[1,-1]), params = params)
  
  # check for hysteresis
  if(soln$behav == "disease-free & endemic equilibria are both locally stable.") {
    # if there's a forward bifurcation, just use simulation to get the realized stable state
    temp <- run_SVI_model(params = par_hyst, init = unlist(sim_df[1,-1]), time = seq(1, 5000, 100))
    inf_df <- data.frame(Group = factor(c("Susceptible", "Vaccinated", "Infected"),
                                      levels = c("Susceptible", "Vaccinated", "Infected")),
                       Value = unlist(temp[50, -1]))
  } else {
    # get the result from the analytical solution
    inf_df <- data.frame(Group = factor(c("Susceptible", "Vaccinated", "Infected"),
                                      levels = c("Susceptible", "Vaccinated", "Infected")),
                         Value = c(soln$S_star, soln$V_star, soln$I_star))
  }
  
  # if vaccinated compartment is fixed at zero for all 
  if (!vaccine) {
    # remove the redundant 0's
    inf_df <- inf_df %>% filter(Group != "Vaccinated")
    inf_df$Group <- droplevels(inf_df$Group)
  }
  
  # plot for infinite time
  inf_plot <- ggplot() +
                geom_hline(data = inf_df, aes(yintercept = Value, color = Group),
                           linewidth = 2.5, alpha = 0.8) +
                scale_x_continuous(limits=c(0.5, 1.5), breaks = c(1), labels = "steady \nstate") +
                scale_colour_viridis_d() +
                scale_y_continuous(limits = c(-0.01, 1.01), expand = c(0, 0)) +
                theme_apps() + theme(text = element_text(size=18),
                                     axis.line.y=element_blank(),
                                     axis.text.y=element_blank(),
                                     axis.title.y=element_blank())
  
  # if specified by user,
  if (!is.na(title_text)){
    # add a title to the main plot
    finite_plot <- finite_plot + labs(title = title_text) +
                    theme(plot.title = element_text(size = 18, face = "bold"))
  }
  
  # put the 2 plots together
  return(wrap_plots(finite_plot, inf_plot, ncol = 2, widths = c(7,1)))
}

# it would be nice to have a function that plots the bifurcation diagram
# plot the bifurcation diagram: R_0 on the x-axis and I^* on the y-axis
# plot_bifurcation <- function(init, params){
# ...
#}

# in lieu of that, here's a function that tells you the final state of the system
get_bifurcation <- function(init, params){
  # extract the parameters for clarity
  beta <- params["beta"]
  gamma <- params["gamma"]
  theta <- params["theta"]
  sigma <- params["sigma"]
  phi <- params["phi"]
  
  # calculate R_0 and R(\phi) as given on pg 188 of Kribs-Zaleta & Velasco-Hernández 2000  
    # R_0 is the reproductive number in the absence of vaccination
  R0 <- (beta / gamma)
  R0 <- unname(R0)
  
  # R(\phi) is R_0 in the presence of vaccination, AKA the vaccine reproduction number
  R0_phi <- R0 * ((theta + sigma*phi) / (theta + phi))
  R0_phi <- unname(R0_phi)
  
  # define B from page 188
  B <- sigma*(beta - gamma) - (theta + sigma*phi)
  
  # summarize model outcome by examining inequalities (see pg 192)
  inequality1 <- (theta + sigma*phi)^2 < gamma * sigma * (1 - sigma) * phi
  inequality2 <- (gamma - ((theta + sigma*phi)/(sigma)) + (2/sigma)*sqrt(gamma*sigma*(1-sigma)*phi)) < beta & beta < gamma*((theta + phi)/(theta + sigma*phi))
  
  if (R0 < 1) {
    behav <- "system always goes to disease-free equilibrium regardless of vaccination."
    I_star <- 0
    V_star <- ifelse(phi + theta > 0, # check for potential undefined values
                     phi / (phi + theta),
                     0)
    
  } else if (phi==0 | is.nan(R0_phi)) { # when theta and phi are 0 then R0_phi is undefined
    # this reduces to the SI model
    if (R0 > 1) {
      behav <- "system always goes to endemic disease equilibrium."
      I_star <- 1 - (1 / R0)
      V_star <- 0
      
    } else { # if R0 is neither smaller nor larger than 1 then it must necessarily be equal to 1
      behav <- "neutral equilibrium: system stays the same as initial."
      I_star <- init["I"]
      V_star <- init["V"]
    }
    
  } else if (all(R0_phi > 1, sigma == 0)) {
    behav <- "system always goes to endemic disease equilibrium."
    # see pg 189
    I_star <- 1 - ((1/R0)*(1 + (phi/theta)))
    V_star <- phi/(R0 * theta)
    
  } else if (R0_phi > 1){
    behav <- "system always goes to endemic disease equilibrium."
    # see pg 191
    I_star <- B/(2 * beta * sigma) # this is broken!!
    V_star <- (phi / beta*sigma) * ((2*beta*sigma - B)/(B + 2*(theta + phi)))
    
  } else if (all(inequality1, inequality2)){
    behav <- "disease-free & endemic equilibria are both locally stable."
    I_star <- c(0,
                B/(2 * beta * sigma))
    V_star <- c(phi / (phi + theta),
                (phi / beta*sigma) * ((2*beta*sigma - B)/(B + 2*(theta + phi))))
    
  } else if (all(R0==1, R0_phi==1)) {
    behav <- "neutral equilibrium: system stays the same as initial."
    I_star <- init["I"]
    V_star <- init["V"]
    
  } else {
    behav <- "system always goes to disease-free equilibrium as a result of vaccination."
    I_star <- 0
    V_star <- ifelse(phi + theta > 0, # check for potential undefined values
                     phi / (phi + theta),
                     0)
  }
  
  I_star <- unname(I_star)
  V_star <- unname(V_star)
  # get S* by subtracting from total population size
  S_star <- rep(1, length(I_star)) - I_star - V_star
  
  output <- list(R0 = R0, R0_phi = R0_phi, I0 = unname(init["I"]),
              S_star = S_star, I_star = I_star, V_star = V_star, behav = behav)
  return(output)
}
```


# R Shiny App

Need to add the hysteresis preset.
But first need to add the bifurcation result raw output.

```{r SVI new_app, echo=FALSE}

# Define UI
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("preset", "Choose preset parameter combo:", 
                  choices = list("SI disease-free" = "SI_DFE",
                                 "SI neutral" = "SI_neutral",
                                 "SI endemic" = "SI_endemic",
                                 "SVI disease-free" = "SVI_DFE",
                                 "SVI vacinated disease-free" = "SVI_vac_DFE",
                                 "SVI neutral" = "SVI_neutral",
                                 "SVI endemic perfect vac" = "SVI_sig0_endemic",
                                 "SVI endemic imperfect vac" = "SVI_imperf_endemic",
                                 "SVI hysteresis" = "SVI_hyst")),
      sliderInput("beta", shiny::HTML("&beta;: Transmission rate"), 
                  min = 0, max = 3, value = 2.27, step = 0.01),
      sliderInput("gamma", shiny::HTML("&gamma;: Recovery rate"),
                  min = 0, max = 3, value = .5, step = 0.01),
      sliderInput("theta", shiny::HTML("&theta;: Vaccine waning rate"),
                  min = 0, max = 3, value = .01, step = 0.01),
      sliderInput("sigma", shiny::HTML("&sigma;: Modifier of transmission rate"),
                  min = 0, max = 1, value = .02, step = 0.02),
      sliderInput("phi", shiny::HTML("&phi;: Vaccination rate"),
                  min = 0, max = 0.5, value = 0.125, step = 0.001),
      sliderInput("I0", label = "Initial fraction of Infected (I)", 
                  min = 0, max = 1, value = 0.1, step = 0.01),
      numericInput("time", label = "Time to simulate", min = 10, value = 100)
    ),
    mainPanel(
      plotOutput("plot_SVI"),
      plotOutput("plot_time")
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  
  observeEvent(input$preset, {
    if (input$preset == "SI_DFE") {
      updateSliderInput(session, "beta", value = 0.7)
      updateSliderInput(session, "gamma", value = 1.1)
      updateSliderInput(session, "theta", value = 0)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0)
    }
    if (input$preset == "SI_neutral") {
      updateSliderInput(session, "beta", value = 0.7)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0)
    }
    if (input$preset == "SI_endemic") {
      updateSliderInput(session, "beta", value = 1.1)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0)
    }
    if (input$preset == "SVI_DFE") {
      updateSliderInput(session, "beta", value = 0.7)
      updateSliderInput(session, "gamma", value = 1.1)
      updateSliderInput(session, "theta", value = 0.1)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0.08)
    }
    if (input$preset == "SVI_vac_DFE") {
      updateSliderInput(session, "beta", value = 1.1)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0.1)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0.08)
    }
    if (input$preset == "SVI_neutral") {
      updateSliderInput(session, "beta", value = 1.3)
      updateSliderInput(session, "gamma", value = 1.3)
      updateSliderInput(session, "theta", value = 0.1)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0.05)
    }
    if (input$preset == "SVI_sig0_endemic") {
      updateSliderInput(session, "beta", value = 1.6)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0.1)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0.05)
    }
    if (input$preset == "SVI_imperf_endemic") {
      updateSliderInput(session, "beta", value = 1.6)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0.18)
      updateSliderInput(session, "sigma", value = 0.06)
      updateSliderInput(session, "phi", value = 0.1)
    }
    if (input$preset == "SVI_hyst") {
      updateSliderInput(session, "beta", value = 1.1)
      updateSliderInput(session, "gamma", value = 0.7)
      updateSliderInput(session, "theta", value = 0)
      updateSliderInput(session, "sigma", value = 0)
      updateSliderInput(session, "phi", value = 0)
    }
  })
  
  V0 <- 0
  
  init_user <- reactive({c(S = 1 - input$I0, V = V0, I = input$I0)})
  init_rareI <- c(S = 1 - 10^-12, V = 0, I = 10^-12)
  init_commonI <- c(S = 10^-4, V = 0, I = 1 - 10^-4)
  
  params <- reactive({c(beta = input$beta, 
                        gamma = input$gamma,
                        theta = input$theta,
                        sigma = input$sigma,
                        phi = input$phi)})
  
  time <- reactive({seq(0, input$time, by = 0.5)})
  
  out_userinit <- reactive({
    data.frame(run_SVI_model(time = time(), params = params(), init = init_user()))
  })
  
  out_rareI <- reactive({
    data.frame(run_SVI_model(time = time(), params = params(), init = init_rareI))
  })
  
  out_commonI <- reactive({
    data.frame(run_SVI_model(time = time(), params = params(), init = init_commonI))
  })
  
  abund_plot_userinit <- reactive({
    plot_SVI_time(sim_df = out_userinit(), params = params())
  })
  
  abund_plot_rare <- reactive({
    plot_SVI_time(sim_df = out_rareI(), params = params(), title_text = "Infections are rare at time = 0:")
  })
  
  abund_plot_common <- reactive({
    plot_SVI_time(sim_df = out_commonI(), params = params(), title_text = "Infections are common at time = 0:")
  })
  
  schematic_plot <- reactive({
    make_rectangles(V_scale = ifelse((V0 == 0) & (input$phi == 0), 0, 1),
                    beta_scale = ifelse(input$beta > 0, (input$beta/3)*2.1 + 0.1, 0),
                    gamma_scale = ifelse(input$gamma > 0, (input$gamma/3)*2.1 + 0.1, 0),
                    phi_scale = ifelse(input$phi > 0, (input$phi/0.5)*2.1 + 0.1, 0),
                    theta_scale = ifelse(input$theta > 0, (input$theta/3)*2.1 + 0.1, 0),
                    sigma_scale = ifelse((input$sigma > 0) & (input$beta > 0),
                                         (input$sigma*input$beta/3)*2.15 + 0.1, 0))
  })
  
  plots_to_render_SVI <- reactive({
    wrap_plots(schematic_plot(), 
               abund_plot_userinit(), 
               nrow = 2, ncol = 1, byrow = FALSE)
  })
  
  time_rare_common <- reactive({
    wrap_plots(abund_plot_rare(), 
               abund_plot_common(), 
               nrow = 2, ncol = 1, byrow = FALSE)
  })
  
  output$plot_SVI <- renderPlot({
    plots_to_render_SVI()
  })
  
  output$plot_time <- renderPlot({
    time_rare_common()
  })
}

# Run the application 
shinyApp(ui = ui, server = server)

```

```{r SVI app, echo=FALSE, eval=FALSE}
## THE FORMAT SHOULD BE CHANGED to make it up to the new standard for R Shiny (i.e., with the ui and server logic as separate functions)

sidebarLayout(
  sidebarPanel(
    
    ### Define some preset parameter combos for debugging ---
    selectInput("preset", "Choose preset parameter combo:", 
                  choices = list("SI disease-free" = "SI_DFE",
                                 "SI neutral" = "SI_neutral",
                                 "SI endemic" = "SI_endemic",
                                 "SVI disease-free" = "SVI_DFE",
                                 "SVI vacinated disease-free" = "SVI_vac_DFE",
                                 "SVI neutral" = "SVI_neutral",
                                 "SVI endemic perfect vac" = "SVI_sig0_endemic",
                                 "SVI endemic imperfect vac" = "SVI_imperf_endemic",
                                 "SVI hysteresis" = "SVI_hyst")),
    
    ### Ask user for parameter values ----
    
    # beta (transmission rate)
    # gamma (recovery rate)
    # theta (vaccine waning rate)
    # sigma (ability of vaccinated to produce new infections)
    # phi (vaccination rate)

    sliderInput("beta", shiny::HTML("&beta;: Transmission rate"), 
                min = 0, max = 3, value = 2.27, step = 0.01), # slider preferences should NOT be hard coded here. Define them as global variables in this code block so can be used for making the rectangle schematic
    
    sliderInput("gamma", shiny::HTML("&gamma;: Recovery rate"),
                min = 0, max = 3, value = .5, step = 0.01),
    
    sliderInput("theta", shiny::HTML("&theta;: Vaccine waning rate"),
                min = 0, max = 3, value = .01, step = 0.01),
    
    sliderInput("sigma", shiny::HTML("&sigma;: Modifier of transmission rate"),
                min = 0, max = 1, value = .02, step = 0.02),
    
    sliderInput("phi", shiny::HTML("&phi;: Vaccination rate"),
                min = 0, max = 0.5, value = 0.125, step = 0.001),
    
    ### Ask user for initial conditions ----
    #numericInput("S0", label = "Initial population size of S", 
    #             min = 0, value = 1 - 0.1),
    #numericInput("V0", label = "Initial population size of V",
    #             min = 0, value = 0),
    sliderInput("I0", label = "Initial fraction of Infected (I)", 
                min = 0, max = 1, value = 0.1, step = 0.01),
    
    ### Ask user for time to simulate ----
    numericInput("time", label = "Time to simulate", min = 10, value = 100)
  ),
  
  # Render plots -----------------
  mainPanel(renderPrint({get_bifurcation(init = init_user(), params = params())}),
            renderPlot({plots_to_render_SVI()}, width = 600, height = 625),
            renderPlot({time_rare_common()}, width = 600, height = 625))
)


# Run the simulation -------------------
# simplify the simulation by setting V0 to zero always
V0 = 0

# Set the initial population sizes
init_user <- reactive({c(S = 1 - input$I0, V = V0, I = input$I0)})
init_rareI <- c(S = 1 - 10^-12, V = 0, I = 10^-12)
init_commonI <- c(S = 10^-4, V = 0, I = 1 - 10^-4)

# Set the parameter values
params <- reactive({c(beta = input$beta, 
                      gamma = input$gamma,
                      theta = input$theta,
                      sigma = input$sigma,
                      phi = input$phi)})
# Time over which to simulate model dynamics
time <- reactive({seq(0, input$time, by = 0.5)})

# Simulate model dynamics 
out_userinit <- reactive({
  data.frame(run_SVI_model(time = time(), params = params(), init = init_user()))
  })

out_rareI <- reactive({
  data.frame(run_SVI_model(time = time(), params = params(), init = init_rareI))
  })

out_commonI <- reactive({
  data.frame(run_SVI_model(time = time(), params = params(), init = init_commonI))
  })

# Make plots -------------------
# Plot abundance through time ----------
abund_plot_userinit <- reactive({
  plot_SVI_time(sim_df = out_userinit(), params = params())
})

abund_plot_rare <- reactive({
  plot_SVI_time(sim_df = out_rareI(), params = params(), title_text = "Infections are rare at time = 0:")
})
abund_plot_common <- reactive({
  plot_SVI_time(sim_df = out_commonI(), params = params(), title_text = "Infections are common at time = 0:")
})


# reactive plot of the model schematic ---
schematic_plot <- reactive({
  make_rectangles(#S_scale = ifelse(input$S0 > 0, input$S0 * 3, 0.05),
                  #I_scale = ifelse(input$I0 > 0, input$I0 * 3, 0.05),
                  #V_scale = input$V0 * 3,
          # remove the V rectangle only when there's NO vaccination and NO vaccinated at time=0
                  V_scale = ifelse((V0 == 0) & (input$phi == 0), 0, 1),
                  beta_scale = ifelse(input$beta > 0, (input$beta/3)*2.1 + 0.1, 0),
                  gamma_scale = ifelse(input$gamma > 0, (input$gamma/3)*2.1 + 0.1, 0),
                  phi_scale = ifelse(input$phi > 0, (input$phi/0.5)*2.1 + 0.1, 0),
                  theta_scale = ifelse(input$theta > 0, (input$theta/3)*2.1 + 0.1, 0),
          # thickness of the m*b arrow depends on BOTH m and b together!
                  sigma_scale = ifelse((input$sigma > 0) & (input$beta > 0),
                                   (input$sigma*input$beta/3)*2.15 + 0.1,
                                   0))
})


# print out the final plots ----

plots_to_render_SVI <- reactive({
  wrap_plots(schematic_plot(), 
             abund_plot_userinit(), 
             nrow = 2, ncol = 1, byrow = FALSE)
})

time_rare_common <- reactive({
  wrap_plots(abund_plot_rare(), 
             abund_plot_common(), 
             nrow = 2, ncol = 1, byrow = FALSE)
})
```

Some notes about how to display this:
- the idea of the interactive schematic plot is interesting but the problem is that lambda is  the force of infection (i.e., it is $\lambda (I)$) so it will be a function of time
- I have another idea of using jitter plot and gganimate to give an idea of how the population is changing over time --> this can express the idea of density and probability of encounter btw different comparments.


Here are some useful parameter combinations that I found by playing around:

```{r param_combos}
# disease-free without vaccination
par_dfe <- c(beta = 0.27, gamma = 0.5, theta = 0.01, sigma = 0.02, phi = 0)

# endemic equilibrium without vaccination


# endemic equilibrium with vaccination


# I think this hysteresis WITHOUT wanning (theta) ?
#test_this <- c(beta = 3.27, gamma = 0.5, theta = 0, sigma = 0.02, phi = 0.01)
# actually, we don't care about this.
# the whole point of why this model is cool is bc addition of the V compartment creates a time delay. ==> There's no way to permanently make individuals resistant to the infection. But it is possible to have a stable disease-free equilibrium bc, through vaccination, it's possible to create conditions where an infected individual is unlikely to encounter any susceptible individuals.
# if we remove the waning, then we have effectively created a resistant (R) compartment in the model.
```

Here's some code for debugging

```{r testings}
# this is a neutral system where R0 == 1 and R0_phi is undefined

test_init <- c(S=0.9, V=0, I=0.1)
test_neut <- c(beta=1.93, gamma=1.93, theta=0, sigma=0, phi=0)

get_bifurcation(init = test_init, params = test_neut)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = test_init, params = test_neut))
plot_SVI_time(sim_df = test_sim, params = test_neut)

init_rareI <- c(S = 1 - 10^-12, V = 0, I = 10^-12)
get_bifurcation(init = init_rareI, params = test_neut)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = init_rareI, params = test_neut))
plot_SVI_time(sim_df = test_sim, params = test_neut)

init_commonI <- c(S = 10^-4, V = 0, I = 1 - 10^-4)
get_bifurcation(init = init_commonI, params = test_neut)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = init_commonI, params = test_neut))
plot_SVI_time(sim_df = test_sim, params = test_neut)
```

```{r testings2}
# this is system with hysteresis
par_hyst <- c(beta = 3.27, gamma = 0.306, theta = 0.01, sigma = 0.02, phi = 0.125)

get_bifurcation(init = test_init, params = par_hyst)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = test_init, params = par_hyst))
plot_SVI_time(sim_df = test_sim, params = par_hyst)

get_bifurcation(init = init_rareI, params = par_hyst)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = init_rareI, params = par_hyst))
plot_SVI_time(sim_df = test_sim, params = par_hyst)

get_bifurcation(init = init_commonI, params = par_hyst)
test_sim <- data.frame(run_SVI_model(time = seq(1, 200, by=2), init = init_commonI, params = par_hyst))
plot_SVI_time(sim_df = test_sim, params = par_hyst)
```


```{to_check}
# why don't the simulated and analytic solutions agree?
to_check1 <- c(beta=1.77, gamma=0.65, theta=0.02, sigma=0.02, phi=0.014)
  # here R0 and R0_phi are both > 1
get_bifurcation(init = c(I=0.3, V=0, S=0.7), params = to_check1)
test_sim <- data.frame(run_SVI_model(time = seq(1, 5000, by=0.1), init = c(I=0.3, V=0, S=0.7), params = to_check1))
plot_SVI_time(sim_df = test_sim, params = to_check1)


# slight discrepency btw simulated and analytic solutions
to_check2 <- c(beta=2.27, gamma=0.5, theta=0.01, sigma=0.02, phi=0.002)
```

# Biological questions & answers to check user understanding

One thing that I learned by playing around with the R Shiny is that it's possible to set m=0 and still have the same kinds of dynamics going on.
